---
title: "acciddasuite"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{acciddasuite}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 7
)
```

# Introduction

The `acciddasuite` package provides tools for building infectious disease forecasts...? This vignette demonstrates how the standard frameworks create and evaluate forecasts using R.

We will aim to demonstrate the basic steps in a forecasting task, as defined by [Hyndman & Athanasopoulos (2021)](https://otexts.com/fpp3/basic-steps.html).

# Statistical Modelling

## Required Packages

The package `tsibble` handles time series data and `fable` [provides forecasting models](https://fable.tidyverts.org/reference/index.html). We also use `fable.prophet` for [Meta's Prophet model](https://facebook.github.io/prophet/).

```{r, setup}
library(tidyverse) # for general data manipulation and plotting
library(pipetime) # time operations in R pipelines (|>)
library(tsibble)
library(fable)
library(fable.prophet) # Prophet model
```

## Loading Data

**Ideally, you would load your own data here**.

For demonstration purposes, we will load data from the [CDC National Health Safety Network](https://data.cdc.gov/Public-Health-Surveillance/Weekly-Hospital-Respiratory-Data-HRD-Metrics-by-Ju/mpgq-jmmr/about_data). The data dictionary is available [here](https://dev.socrata.com/foundry/data.cdc.gov/mpgq-jmmr) and provides [instructions](https://dev.socrata.com/foundry/data.cdc.gov/mpgq-jmmr#:~:text=from%20Prior%20Week-,Code%20Snippets) on how to load the data in R using the `RSocrata` package.

We extract data for California, Texas, and New York starting from March 1, 2022.

```{r, epidatr}
library(RSocrata)

df <- read.socrata(
  "https://data.cdc.gov/api/v3/views/mpgq-jmmr"
) |>
  filter(jurisdiction %in% c("CA", "TX", "NY")) |>
  transmute(
    date = as.Date(weekendingdate),
    state = jurisdiction,
    value = totalconffluhosppats
  ) |>
  filter(date >= as.Date("2022-03-01")) |>
  tsibble::as_tsibble(index = date, key = state)

df |>
  autoplot(value) +
  labs(
    title = "Weekly Flu Hospitalisations",
    y = "Total Confirmed Patients",
    x = "Week Ending"
  ) +
  theme_classic()

```

## Time Series Cross-Validation

To evaluate predictive performance, we employ time series cross-validation (rolling forecasting origin). We fit models using only data available up to a specific cutoff point, then forecast 4 weeks ahead.\
Initial training window: 50 weeks. Step size: Advance the cutoff by 4 weeks.\
Forecast horizon ($h$): 4 weeks.\

```{r, stretch}
ts <- df |>
  stretch_tsibble(.init = 50, .step = 4)
```

## Model Specification and Forecasting

We compare three forecasting approaches on the folded data: **Naïve (Random Walk RW)**: A baseline model carrying forward the last observation. **ETS**: Exponential Smoothing state space model (automatically selected). **ARIMA**: Auto-Regressive Integrated Moving Average model (automatically selected). The code below fits these models to every window defined in `ts` and generates a 4-week forecast for each. The below can take up to 3min on a standard laptop.

```{r, models}
fcast <- ts |>
  model(
    RW = RW(value),
    ETS = ETS(value),
    ARIMA = ARIMA(value),
    THETA = THETA(value),
    PROPHET = fable.prophet::prophet(value)
  ) |>
  time_pipe("model_fit") |>
  forecast(h = 4) |>
  time_pipe("forecasting")
```

## Model Evaluation

### By State

We assess which model performed best on average for each state. We calculate the RMSE (Root Mean Squared Error) across all rolling windows, grouped by state and model.

```{r, state-evaluation}
model_colours <- c(
  "RW" = "#2E86C1",
  "ETS" = "#28B463",
  "ARIMA" = "#CB4335",
  "THETA" = "#AF7AC5",
  "PROPHET" = "#F39C12"
)

state_accuracy <- fcast |>
  fabletools::accuracy(df, by = c("state", ".model")) |>
  select(state, .model, RMSE) |>
  (\(x) {
    bind_rows(
      x,
      x |>
        group_by(.model) |>
        summarise(
          state = "ALL",
          RMSE = mean(RMSE, na.rm = TRUE),
          .groups = "drop"
        )
    )
  })()

state_accuracy |>
  group_by(state) |>
  mutate(.model = fct_reorder(.model, RMSE)) |>
  ggplot(aes(x = .model, y = RMSE, fill = .model)) +
  facet_grid(rows = vars(state)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = model_colours) +
  theme_classic() +
  theme(legend.position = "none")

```

### Over Time

Aggregated metrics can hide temporal instability (*i.e.* a model failing during a specific outbreak peak). Here we compute the RMSE for rolling window (fold) to visualise how model performance evolves over the epidemic curve.

```{r, time-evaluation}
fold_accuracy <- fcast |>
  fabletools::accuracy(df, by = c(".model", "state", ".id"))

#Map .id back to the specific cutoff date for plotting
fold_dates <- ts |>
  as_tibble() |>
  group_by(.id) |>
  summarise(cutoff_date = max(date))


left_join(fold_accuracy, fold_dates, by = ".id") |>
  ggplot(aes(x = cutoff_date, y = RMSE, color = .model)) +
  geom_line() +
  facet_grid(rows = vars(state)) +
  scale_color_manual(values = model_colours) +
  theme_classic()
```

## Visualising Forecasts with Intervals

We can also visualise the rolling forecasts along with their 95% prediction intervals.

```{r, forecast-plot}
fcast |>
  hilo(level = 95) |>
  mutate(
    lower = `95%`$lower,
    upper = `95%`$upper
  ) |>
  ggplot(aes(x = date, group = interaction(.id, .model))) +
  geom_ribbon(
    aes(ymin = lower, ymax = upper, fill = .model),
    alpha = 0.2
  ) +
  geom_line(aes(y = .mean, color = .model), linewidth = 0.5) +
  geom_line(
    data = df,
    aes(x = date, y = value, group = state),
    color = "black",
    linewidth = 0.5,
    inherit.aes = FALSE
  ) +
  facet_grid(rows = vars(state)) +
  labs(
    title = "Rolling Forecasts with 95% Confidence Intervals",
    subtitle = "Black line = Observed Data; Coloured bands = Retrospective Forecasts",
    y = "Hospitalisations",
    x = "Date"
  ) +
  theme_classic() +
  theme(legend.position = "bottom")
```

## Forecasting Future Values

Finally, we fit the best-performing model (based on prior evaluation) to the entire dataset and generate forecasts for the next 4 weeks. We also show NNETAR as an alternative model but it takes a lot of time to fit (hence why we didn't use it in the previous steps). This can take up to 1min on a standard laptop.

```{r, final-forecast}
finalcast <- df |>
  model(
    THETA = THETA(value),
    NNETAR = NNETAR(value, n_networks = 1)
  ) |>
  forecast(h = 10) |> #10 weeks ahead
  time_pipe("final_forecast")

finalcast |>
  autoplot(df) +
  theme_classic() +
  facet_grid(rows = vars(state))
```

# Epidemic Models

Epidemiologists are often interested in characterising the dynamics of infectious diseases, i.e. the mechanisms that drive transmission.\
Rather than relying solely on statistical models, epidemic models can be used to explicitly represent the mechanisms of transmission. Here we demonstrate how to estimate the time-varying reproduction number (Rt) and use it to generate short-term forecasts.
Rt represents the average number of secondary infections generated by a primary case at time t. When Rt > 1, the epidemic is growing; when Rt < 1, it is declining.

## Rt Estimation

### Required Packages

```{r, rt-setup}
library(EpiEstim)
#library(EpiNow2)
library(projections)
library(incidence)
```

### Estimating Rt

```{r, rt-estimation}
states_Rt <- tibble(
  state = c("CA", "TX", "NY"),
  data = map(
    state,
    ~ as_tibble(df) |>
      filter(state == .x) |>
      select(date, value) |>
      rename(I = value)
  ),
  R = map(
    data,
    ~ EpiEstim::estimate_R(
      incid = .x$I,
      method = "parametric_si",
      config = EpiEstim::make_config(
        list(
          mean_si = 3.96,
          std_si = 4.75
        )
      )
    )
  )
)

# projections for California
CA <- states_Rt |>
  filter(state == "CA")
CA

incidence_obj <- pluck(CA, "data", 1) |>
  mutate(I = round(I)) |> #we convert average to counts
  uncount(weights = I) |>
  pluck("date") |>
  incidence()

proj <- project(
  x = incidence_obj,
  R = pluck(CA, "R", 1, "R", "Median(R)"), # Rt distribution
  si = pluck(CA, "R", 1, "si_distr")[-1], # si starts at day 1
  n_sim = 100, # number of trajectories to simulate
  n_days = 30, # number of days to make projections over
  R_fix_within = FALSE # keep the same value of R every day (TRUE/FALSE)
)

cum_proj <- proj |> cumulate()
plot(incidence_obj) |>
  add_projections(cum_proj, boxplots = FALSE, quantiles = c(0.25, 0.75)) +
  scale_x_date(
    limits = c(max(incidence_obj$dates) - 365, max(incidence_obj$dates) + 30)
  ) +
  theme_classic()

```